The effect of linking assumptions and number of response options on inferred scalar implicature rate

### 2018-05-02 writeup.Rmd
minor gd edits

We begin by reporting an experiment that addresses the following question: do the number of response options provided in a truth-value judgment task and the way that responses are grouped into pragmatic ('SI') and literal ('no SI') change inferences about scalar implicature rates? Note that this way of asking the question presupposes two things: first, that whatever participants are doing in a truth-value judgment task, the behavioral measure can be interpreted as providing a measure of **interpretation**. And second, that listeners either do or do not compute an implicature on any given occasion. In the Discussion we will discuss both of these issues. First, following @Degen2014, we will offer some remarks on why truth-value judgment tasks are better thought of as measuring participants' estimates of speakers' **production** probabilities. This will suggest a completely different class of linking functions. And second, we discuss an alternative conception of scalar implicature as a probabilistic phenomeonen, a view that has recently rose to prominence in the subfield of probabilistic pragmatics. This alternative conception of scalar implicature, we argue, affords developing and testing quantitative linking functions in a rigorous and motivated way.

@@

# Discussion
=======
On the traditional view of the link between implicature and behavior in sentence verification tasks, scalar implicature is conceptualized as a binary, categorical affair - that is, an implicature is either ‘calculated’ or it isn’t, and the behavioral reflexes of this categorical interpretation process should be straightforwardly observed in experimental paradigms. This assumption has concerning implications for how we must approach analysis of variation in behavior on a truth value judgment task; for example, why did the majority of respondents in the binary condition of our experiment answer “Right” to an utterance of cat or dog when the card had both a cat and a dog on it?

To explain the data on the traditional view, are forced to say that a) not all participants calculated the implicature; or that b) some participants who calculated the implicature did not choose the anticipated response (i.e. “Wrong”) due to some other cognitive reflex which ‘overrode’ the implicature; or some mixture of (b) and (c). We might similarly posit that one or both of these factors underlie the variation in the ternary, quatenary, and quinary conditions (e.g. why were participants roughly split between “Right” and “Kind of right” when the utterance was cat or dog and the card had a cat and a dog?). However, the best we can hope for on this approach is an analysis which traces the general qualitative patterns in the data.

We contrast the above view of implicature and its behavioral reflexes with an alternative linking hypothesis which assumes that participants' behavior can be represented using the model of a soft-optimal pragmatic speaker in the RSA framework. This alternative linking hypothesis contrasts with the traditional view in it is rooted in a quantitative formalization of pragmatic competence which provides us a continuous measure of pragmatic reasoning. Recall that in RSA, pragmatically competent listeners are modeled as a continuous probabilistic distribution of possible meanings given an utterance which that listener hears. The probability with which this listener $L_1$ ascribes a meaning s to an utterance u depends upon a prior probability distribution of potential states of the world $P_w$, and upon reasoning about the communicative behavior of a speaker $S_1$. $S_1$ in turn is modeled as a continuous probabilistic distribution of possible utterances given an intended state of affairs the speaker intends to communicate. This distribution is sensitive to a rationality parameter $\alpha$, the production cost $C$ of potential utterances, and a representation of a literal listener $L_0$ whose interpretation of an utterance is in turn a function of that utterance's truth conditional content $[[u]](s)$ and her prior beliefs about the state of the world $P_w(s)$.

@@

In this framework, individuals never categorically draw (or fail to draw) pragmatic inferences about the utterances they hear. For example, exclusivity readings of disjunction or are represented in RSA as relatively low conditional probability of a conjunctive meaning on the $P_L$ distribution, given an utterance of or. Thus, it is not even possible to talk about 'rate' of implicature calculation in the RSA framework. The upshot, as we show below, is that this view of pragmatic competence does allow us to talk explicitly and quantitatively about rates of observed behavior in sentence verification tasks.

First, following Degen & Goodman (2014), we proceed on the assumption that behavior on sentence verification tasks, such as truth value judgment tasks, is best modeled as a function of an individual’s mental representation of a cooperative interlocutor ($S_1$ in the language of RSA) rather than of a pragmatic listener who interprets utterances ($P_{L_1}$). In their paper, Degen & Goodman argue that sentence verification tasks are relatively more sensitive to contextual manipulations (such as manipulation of the Question Under Discussion) than are sentence interpretation tasks, and that this follows if sentence interpretation tasks - but not sentence verification tasks - require an additional layer of counterfactual reasoning about the intentions of a cooperative speaker.

A main desideratum of a behavioral linking hypothesis given the RSA view of pragmatic competence is to transform continuous probability distributions into categorical outputs (e.g. responses of “Right”/”Wrong” in the case of the binary condition of our experiment). For a given utterance u and an intended communicated meaning w, $S_1$(u | w) outputs a conditional probability of u given w. For example, in the binary condition of our experiment where a participant evaluated cat or dog when there were both animals on the card, the participant has access to the mental representation of $S_1$ and hence to the $S_1$ conditional probability of hearing the utterance cat or dog given a dog and cat card: $S_1$(cat or dog | cat and dog). According to the linking hypothesis advanced here, the participant provides a particular response to u if the RSA speaker probability of u lies within a particular probability interval, given an observed state of the world (i.e. the configuration of animals on the card in our experiment). We model a responder, R, who in the binary condition responds “Right” to an utterance u in world w just in case $S_1$(u | w) exceeds some probability threshold $\theta$:

@@

= “Right” iff $S_1$(u | w) $>$ $\theta$

@@

In the experiment conditions where there are more than two choices, we model the range of possible behavioral responses for R with the introduction of intermediate probability thresholds. For example, in the ternary condition, R(u, w, $\theta_1$ , $\theta_2$) is “Right” iff $S_1$(u | w) > $\theta_1$ and “Neither” iff $\theta_1$ > $S_1$(u | w) > $\theta_2$. To fully generalize the model to our five experimental conditions, we say that R takes as its input an utterance u, a world state w, and a number of threshold variables dependent on a variable c, corresponding to the experimental condition in which the participant finds herself (e.g. the range of possible responses available to R).

>>>>

We begin by reporting an experiment that addresses the following question: do the number of response options provided in a truth-value judgment task and the way that responses are grouped into pragmatic ('SI') and literal ('no SI') change inferences about scalar implicature rates? Note that this way of asking the question presupposes two things: first, that whatever participants are doing in a truth-value judgment task, the behavioral measure can be interpreted as providing a measure of **interpretation**. And second, that listeners either do or do not compute an implicature on any given occasion. In the General Discussion we will discuss both of these issues. First, following @Degen2014, we will offer some remarks on why truth-value judgment tasks are better thought of as measuring participants' estimates of speakers' **production** probabilities. This will suggest a completely different class of linking functions. And second, we discuss an alternative conception of scalar implicature as a probabilistic phenomeonen, a view that has recently rose to prominence in the subfield of probabilistic pragmatics. This alternative conception of scalar implicature, we argue, affords developing and testing quantitative linking functions in a rigorous and motivated way.

@@

# General Discussion
=======

On the traditional view of the link between implicature and behavior in sentence verification tasks, scalar implicature is conceptualized as a binary, categorical affair - that is, an implicature is either ‘calculated’ or it isn’t, and the behavioral reflexes of this categorical interpretation process should be straightforwardly observed in experimental paradigms. This assumption raises concerns for analyzing variation in behavior on a truth value judgment task; for example, why did the majority of respondents in the binary condition of our experiment answer “Right” to an utterance of the underinformative "There is a cat or dog" when the card had both a cat and a dog on it?

To explain the data on the traditional view, we are forced to say that a) not all participants calculated the implicature; or that b) some participants who calculated the implicature did not choose the anticipated (i.e., "Wrong") response due to some other cognitive process which overrode the 'correct' implicature behavior; or some mixture of (a) and (b). We might similarly posit that one or both of these factors underlie the variation in the ternary, quaternary, and quinary conditions. However, the best we can hope for on this approach is an analysis which traces the general qualitative patterns in the data. **XXX Brandon, can you be more specific about what you mean by 'tracing a general qualitative pattern'?**

We contrast the above view of implicature and its behavioral reflexes with an alternative linking hypothesis. Recent developments in the field of probabilistic pragmatics have demonstrated that pragmatic production and comprehension can be captured within the Rational Speech Act (RSA) framework, which treats listeners and speakers as agents that reason about each other and choose interpretations and utterances, respectively, that soft-maximize utility [@Frank2012; @Goodman2013; @Goodman2016; @Franke2016; @Qing2014; @Degen2013; @DegenTG2015; @Bergen2016; @Kao2014; @Scontras2017]. We take inspiration from this approach and treat participants' behavior in our experimental tasks as the result of a soft-optimal pragmatic speaker in the RSA framework. This alternative linking hypothesis contrasts with the traditional view in that it is rooted in a quantitative formalization of pragmatic competence which provides us a continuous measure of pragmatic reasoning. In RSA, pragmatically competent listeners are modeled as a continuous probabilistic distribution over possible meanings (states of the world) given an utterance which that listener observes. The probability with which this listener $L_1$ ascribes a meaning $s$ to an utterance $u$ depends upon a prior probability distribution of potential states of the world $P_w$, and upon reasoning about the communicative behavior of a speaker $S_1$. $S_1$ in turn is modeled as a continuous probabilistic distribution over possible utterances given an intended state of the world the speaker intends to communicate. This distribution is sensitive to a rationality parameter $\alpha$, the production cost $C$ of potential utterances, and the informativeness of the utterance, quantified via a representation of a literal listener $L_0$ whose interpretation of an utterance is in turn a function of that utterance's truth conditional content $[[u]](s)$ and her prior beliefs about the state of the world $P_w(s)$.

@@

In this framework, individuals never categorically draw (or fail to draw) pragmatic inferences about the utterances they hear. For example, exclusivity readings of disjunction are represented in RSA as relatively lower conditional probability of a conjunctive meaning on the $P_L$ distribution given an utterance of "or", compared to its prior probability. Thus, it is not even possible **XXX or at least not without auxiliary assumptions** to talk about 'rate' of implicature calculation in the RSA framework. The upshot, as we show below, is that this view of pragmatic competence does allow us to talk explicitly and quantitatively about rates of observed behavior in sentence verification tasks.

First, following @Degen2014, we proceed on the assumption that behavior on sentence verification tasks, such as truth value judgment tasks, is best modeled as a function of an individual’s mental representation of a cooperative interlocutor ($S_1$ in the language of RSA) rather than of a pragmatic listener who interprets utterances ($P_{L_1}$). In their paper, Degen & Goodman argue that sentence verification tasks are relatively more sensitive to contextual manipulations (such as manipulation of the Question Under Discussion) than are sentence interpretation tasks, and that this follows if sentence interpretation tasks - but not sentence verification tasks - require an additional layer of counterfactual reasoning about the intentions of a cooperative speaker.

A main desideratum of a behavioral linking hypothesis given the RSA view of pragmatic competence is to transform continuous probability distributions into categorical outputs (e.g. responses of “Right”/”Wrong” in the case of the binary condition of our experiment). For a given utterance $u$ and an intended communicated meaning $s$, $S_1$(u | s) outputs a conditional probability of $u$ given $s$. For example, in the binary condition of our experiment where a participant evaluated "There is a cat or a dog" when there were both animals on the card, the participant has access to the mental representation of $S_1$ and hence to the $S_1$ conditional probability of producing the utterance "cat or dog"" given a dog and cat card: $S_1$("cat or dog" | cat and dog). According to the linking hypothesis advanced here, the participant provides a particular response to $u$ if the RSA speaker probability of $u$ lies within a particular probability interval, given an observed state of the world (i.e. the configuration of animals on the card in our experiment). We model a responder, $R$, who in the binary condition responds “Right” to an utterance $u$ in world $s$ just in case $S_1(u | s)$ exceeds some probability threshold $\theta$:

@@

= “Right” iff $S_1$(u | s) $>$ $\theta$

@@

In the experiment conditions where there are more than two choices, we model the range of possible behavioral responses for $R$ with the introduction of intermediate probability thresholds. For example, in the ternary condition, R(u, s, $\theta_1$ , $\theta_2$) is “Right” iff $S_1$(u | s) > $\theta_1$ and “Neither” iff $\theta_1$ > $S_1$(u | s) > $\theta_2$. To fully generalize the model to our five experimental conditions, we say that $R$ takes as its input an utterance $u$, a world state $s$, and a number of threshold variables dependent on a variable $c$, corresponding to the experimental condition in which the participant finds herself (e.g. the range of possible responses available to $R$).
